Project 3: Variant calling in Burbot (*Lota lota*) using `freebayes` and `mpileup`

### 2021-03-19 manuscript.Rmd
wrote a discussion, of sorts

# Project 3: Variant calling in Burbot (*Lota lota*) using `freebayes` and `mpileup`

@@

In the current work I compared variant calling by two tools: `bcftools mpileup/call` (previously `samtools`) and `freebayes`. The results were compared in terms of the number of SNPs, the overlap between sets, the depth at each position, and their minor allele frequencies. Both `bcftools` and `freebayes` apply Bayesian models to estimate genotype posterior probabilities, though they differ in how this is implemented. `bcftools` uses a two step process where `mpileup` first creates the 'pileup' of all reads at variant sites, and then `call` is used to infer genotypes at these [@li2011, @danecek2021]. This implementation has the advantage of being very fast. `freebayes` a uses different approach to inferring haplotypes in that it does not use the alignment but the full reads [@garrison2012, @howtof]. Previous comparisons of variant callers have often found low concordance between tools and issues with sensitivity [@orawe2013, @cornish2015]. However, others have reported high concordance and accuracy in their evaluation [@hwang2015]; the best method is likely dependent on nature of the sequencing data: (whole-genome or exome, RAD-seq, RNA-seq, *etc*.), the depth of coverage, and the type of variants of interest (*eg.* SNPs or indels).

@@

I used data from project \#2 in this work, specifically the reads mapped to the burbot reference with the `bowtie2` aligner. I used `bcftools mpileup` and `GATK HaplotypeCaller` to call variants for the ten individuals using these alignments. I first needed to convert my `.sam` files to `.bam` format and index them, as well as adding read groups using `picard` (necessary for `freebayes`).

@@

I ran `bcftools` variant calling for the 10 .sorted.bam files. For `mpileup`, I added arguments to provide annotations for read depth and allelic depth (-a DP,AD) and to specify Illumina as the sequencing platform (-P). For `call`, I provided arguments to specify the use of the multi-allelic caller, 'variants only', to skip indels, and to provide the genotype quality scores (GQ).

@@

I called SNPs using `freebayes` with default settings; annotations for genotype quality were requested with `--genotype-qualities`. Output was limited to SNPs by piping to `vcffilter`.

@@

I used the `vcftools` library to filter variant calls. Biallelic SNPs (`--min-alleles 2 --max-alleles 2 --remove-indels`) were retained if they had a quality score of at least 20 (`--minQ 30`). Genotypes were retained at these sites if they had genotyping quality score of at least 10 (`--minGQ 10`). I found the vast majority of variants to be shared among all 10 individuals, so I applied filters for the maximum and minimum minor allele frequencies, such that these would not be either zero or one (`–maf 0.01 –max-maf 0.99`). Any calls that did not receive a 'PASS' flag were discarded (`--remove-filtered-all`).

@@

```{r overlap_in_sets, engine = 'bash', eval = FALSE, echo = TRUE}
# prepare files for isec by zipping and re-indexing
bgzip mpileup.filter.recode.vcf
bcftools index mpileup.filter.recode.vcf.gz
bgzip freebayes.filter.recode.vcf
bcftools index freebayes.filter.recode.vcf.gz
# Compute intersection of 2 files with isec
bcftools isec -p isec mpileup.filter.recode.vcf.gz freebayes.filter.recode.vcf.gz
```

<!-- # Get overlap data from isec into tsv format -->

<!-- > Overlap.vals -->

<!-- for file in isec/*.vcf -->

<!-- do  -->

<!--   bcftools stats $file| grep "number of SNPs:" | cut -f4 >> Overlap.vals -->

<!-- done -->

<!-- echo $'unique to mpileup\nunique to freebayes\nshared\nshared' > Overlap.names -->

<!-- paste  Overlap.names Overlap.vals > Overlap.tsv -->

<!-- sed -i '1s/^/SNPs\tCount\n/' Overlap.tsv -->

Using `vcftools`, I computed the following:

-   Allele frequency at each loci across individuals

```{r vcf_freq, engine = 'bash', eval = FALSE, echo = TRUE}
vcftools --gzvcf mpileup.filter.recode.vcf.gz --freq \
  --out mpileup.filter
vcftools --gzvcf freebayes.filter.recode.vcf.gz --freq --out freebayes.filter
```

-   Sum depth at each locus across individuals

```{r vcf_site_depth, engine = 'bash', eval = FALSE, echo = TRUE}
vcftools --gzvcf mpileup.filter.recode.vcf.gz --site-depth --out mpileup
vcftools --gzvcf freebayes.filter.recode.vcf.gz --site-depth --out freebayes
```

-   Mean depth for each individual across all loci

```{r vcf_indv_depth, engine = 'bash', eval = FALSE, echo = TRUE}
vcftools --gzvcf mpileup.filter.recode.vcf.gz --depth --out mpileup
vcftools --gzvcf freebayes.filter.recode.vcf.gz --depth --out freebayes
```

<!-- ## Relatedness -->

<!-- vcftools --gzvcf mpileup.filter.recode.vcf.gz --relatedness --out mpileup -->

<!-- vcftools --gzvcf freebayes.filter.recode.vcf.gz --relatedness --out freebayes -->

<!-- ## Convert to (0,1,2) format -->

<!-- vcftools --gzvcf mpileup.filter.recode.vcf.gz --012 --out mpileup -->

<!-- vcftools --gzvcf freebayes.filter.recode.vcf.gz --012 --out freebayes -->

<!-- quality, as shown in class, using what you think is a reasonable threshold. Compare the resulting vcf files in several ways. First, *compare the number of SNPs* and the *overlap of these SNPs across files*. -->

<!-- Then, plot the *minor allele frequency* and depth of all SNPs (summed across individuals), and discuss the variation across loci. Compare across vcfs produced by different variant callers. -->

```{=html}
<!-- # get data down from cedar in local terminal:
 scp jmoggrid@cedar.computecanada.ca:/scratch/jmoggrid/Project3/Overlap.tsv .
 scp jmoggrid@cedar.computecanada.ca:/scratch/jmoggrid/Project3/*.frq . 
 scp jmoggrid@cedar.computecanada.ca:/scratch/jmoggrid/Project3/*.idepth . 
 scp jmoggrid@cedar.computecanada.ca:/scratch/jmoggrid/Project3/*.ldepth .  -->
```

@@

I created figures and summary tables from the data generated above using the `R` language.

@@

# Overlap counts table; add
Overlap <- read_delim('./data/Overlap.tsv', '\t') %>%  
  distinct() %>% 
  bind_rows(
    tibble(SNPs = c('mpileup', 'freebayes'), 
           Count = c(.[1,2][[1]] + .[3,2][[1]], 
                     .[2,2][[1]] + .[3,2][[1]]))) %>%  
  arrange(desc(Count))

@@

vcf_df <- SiteDepth %>%
  rename(chrom = chr) %>% 
  left_join(AlleleFreq, by = c('chrom', 'pos', 'caller'))
# glimpse(vcf_df)

cnames <- c('id', 'quality', 'n_snps', 'ts', 'tv', 'indels')
qual_df <- 
  read_tsv("./data/freebayes.quality.stats",
                 skip = 1L, col_names = cnames) %>% 
  mutate(caller = "freebayes") %>% 
  bind_rows(
    read_tsv("./data/mpileup.quality.stats",
                 skip = 1L, col_names = cnames) %>% 
  mutate(caller = "mpileup")
  )

```

```{r figs, echo=F}
library(ggbeeswarm)
library(patchwork)
library(rcartocolor)
library(gridExtra)
library(grid)
# Allele frequency histogram (alleles stacked)
fig_allelefreq <- 
  ggplot(AlleleFreq) +
  geom_histogram(aes(x = maf), bins = 20, alpha = 0.8) +
  facet_wrap(caller~., scales = 'free_y') +
  labs(x = 'minor allele frequency', y='loci') +
  theme_bw()

@@

# The distribution of mean depth per site within individuals
set.seed(45)
fig_indvdepth <-
  ggplot(IndvDepth, aes(y = mean_depth, x = caller)) +
  geom_violin(alpha = 0.1, draw_quantiles = c(.25, .5, .75), trim = F) +
  geom_quasirandom(aes(color = individual), size = 3, width = 0.3) +
  scale_color_carto_d(palette = 2) +
  guides(colour = guide_legend(override.aes = list(size = 4, pch = 16))) +
  facet_wrap( ~ caller, scales = 'free_x') +
  labs(y = 'mean depth per locus', x = '', color = 'Individual') +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = 'right'
  )

# Allele depth over maf, coloured by n chromosomes
fig_alleledepth_over_maf <- 
  ggplot(vcf_df, aes(x=maf, y=allele_depth, color = n_chr)) +
  geom_point(size = 0.5, alpha = 0.7) +
  scale_y_log10() +
  facet_wrap(~caller) +
  scale_color_viridis_c() +
  labs(x = "Minor allele frequency",
       y = "Allele depth", 
       color="Number of\nchromosomes") +
  theme_bw()
```

@@

The distribution minor allele frequencies of SNPs from `mpileup` shows a U-shape with peaks at 0.05 and 0.5, whereas those from `freebayes` peak at 0.05 and are otherwise evenly distributed (fig.1A). The sum depth per locus across all individuals is generally in the range of 50-200 reads, with both callers showing similar distributions (fig.1B). The mean depth for `mpileup` was 173.7 reads and for `freebayes` 162.2 reads. Both callers have some extreme outliers in terms of site depth: the maximum depth for `freebayes` was 16,162 reads, while for `mpileup` it was only 2522 reads. The minimum site depth was 5 and 3 for `freebayes` and `mpileup` respectively and `mpileup` has a greater proportion of low-depth sites.

@@

```{r table2_overlap, echo=F}
Overlap %>% pander::pander(caption = "Total, shared, and unique SNPs called by mpileup and freebayes (after filtering).")
```

```{r fig1, echo=F, fig.align = 'center', fig.width=7, fig.height=3, fig.cap="(A) Distribution of minor allele frequency; SNPs with frequency = 0 or 1 were omitted. (B) Distribution of depth per locus summed across all individuals (note log-scaled x-axis)."}

(fig_allelefreq + fig_sitedepth) &
plot_layout(guides = 'keep') &
plot_annotation(tag_levels = 'A')
```

```{=html}
<!-- Zeggini, E., & Morris, A. (2015). Assessing Rare Variation in Complex Traits Design and Analysis of Genetic Studies (1st ed. 2015.). Springer 
New York. <https://doi.org/10.1007/978-1-4939-2824-8> -->
```
```{r engine = 'bash', eval=FALSE}
## quality scores...
grep -E "Quality" mpileup.filter.recode.stats | cut -f1 --complement > mpileup.quality.stats
grep ^QUAL mpileup.filter.recode.stats | cut -f1 --complement >> mpileup.quality.stats 
grep -E "Quality" freebayes.filter.recode.stats | cut -f1 --complement > freebayes.quality.stats
grep ^QUAL freebayes.filter.recode.stats | cut -f1 --complement >> freebayes.quality.stats 
```

```{r echo = F}
fig_quality_dist <- qual_df %>% 
  mutate(QUAL = round(quality/25)*25) %>% 
  group_by(QUAL, caller) %>% 
  summarise(n_snps = sum(n_snps),
            ts  = sum(ts),
            tv = sum(tv)) %>% 
  pivot_longer(cols = c(ts,tv)) %>% 
  ggplot(aes(x = QUAL, y = value, fill = name)) +
  geom_col(width = 19) +
  facet_wrap(~caller) +
  scale_fill_carto_d(palette = 2) +
  labs(x="quality score", y = "loci", fill = NULL) +
  scale_y_log10() +
  theme_bw()

```

@@

<!-- Sites with low quality were discarded -->

<!-- Kept all allele freq, even rare ones ... rare ones are cool because reference... depends of system of interest though ... -->

<!-- Sites with extremely large depth could represent paralogous regions? -->

Comparison of callers: Hwang2015; Cornish2015

-   Reflect on results

-   Concurrence with other findings

    -   If not, why not (data...)

-   Characteristics of the data affect variant calls?

    -   Reference was highly fragmented, though \~85 % reads mapped to the reference in all individuals (in earlier work)

    -

-   Other features of the VCF that are notable ? The mean read depth across all sites is larger for all individuals in the `freebayes` SNPs (fig. 2). Perhaps this is because freebayes is less tolerant of low-coverage sites in it's estimation of genotype likelihood/ quality / ....

-   What other filters:

    -   Depth.... high coverage can indicate a paralogous region (duplicated)

    -   Might be interested in triallelic snps or indels, depending on the study...

    -   If looking for rare variants we would perhaps want to not use filters and only look at positions that are relevant to the process or disease of interest [@cornish2015].

Both `freebayes` and `mpileup` apply Bayes' theorem in their algorithms for SNP calling.

mpileup considers the bases at a variant position

freebayes uses a window approach that differs from other bayesian callers which treat the alignment similarly...

cornish

-   callers have high positive predictive value (TP/TP+FP) but low sensitivity (TP/P).

-   Freebayes identified more raw SNPs (far more false positive) in the human exome data; but mpileup and freebayes had similar counts after filtering.

-   Freebayes did not perform as well as mpileup.

# Supplemental Figures

```{r}
fig_alleledepth_over_maf
```

```{r fig2, echo=F, fig.align = 'center', fig.width=5, fig.height=2.9, fig.cap="Mean depth per locus for each individual."}
fig_indvdepth
```

```{r fig3, echo=F, fig.align = 'center', fig.width=5, fig.height=2.9, fig.cap="Mean depth per locus for each individual."}
fig_alleledepth_over_maf
```

```{r}

```

>>>>

# Project 3: Comparison of variant calling by `freebayes` and `mpileup` from high-throughput sequencing of Burbot (*Lota lota*)

@@

In the current work I compared variant calling by two tools: `bcftools mpileup/call` (previously `samtools`) and `freebayes`. The results were compared in terms of the number of SNPs, the overlap between sets, the depth at each position, and their minor allele frequencies. Both `bcftools` and `freebayes` apply Bayesian models to estimate genotype posterior probabilities, though they differ in how this is implemented. `bcftools` uses a two step process where `mpileup` first creates the 'pileup' of all reads at variant sites, and then `call` is used to infer genotypes at these [@li2011, @danecek2021]. This implementation has the advantage of being very fast. `freebayes` a uses different approach to inferring haplotypes in that it does not use the alignment but the full reads [@garrison2012; @howtof]. Previous comparisons of variant callers have often found low concordance between tools and issues with sensitivity [@orawe2013, @cornish2015]. However, others have reported high concordance and accuracy in their evaluation [@hwang2015]; the best method is likely dependent on nature of the sequencing data: (whole-genome or exome, RAD-seq, RNA-seq, *etc*.), the depth of coverage, and the type of variants of interest (*eg.* SNPs or indels).

@@

I used data from project \#2 in this work, specifically the reads mapped to the burbot reference with the `bowtie2` aligner. First, I needed to convert my `.sam` files to `.bam` format and index them, as well as adding read groups using `picard` (necessary for `freebayes`):

@@

I used `bcftools mpileup/call` (referred to as `mpileup`) and `freebayes` to call variants from the ten burbot alignments. For `mpileup`, I added arguments to include annotations for read depth and allelic depth (-a DP,AD) and to specify Illumina as the sequencing platform (-P). For `call`, I specified the use of the newer multi-allelic caller (m), 'variants only' (v), to skip indels, and to provide the genotype quality scores (GQ). A summary of the raw SNP calls was obtained using `bcftools stats`.

@@

I called SNPs using `freebayes` with default settings except for the flag to provide annotations for genotype quality. Output was limited to SNPs with `vcffilter`.

@@

I used the `vcftools` library to filter variant calls. Biallelic SNPs (`--min-alleles 2 --max-alleles 2 --remove-indels`) were retained if they had a quality score of at least 20 (`--minQ 20`). Genotypes were retained at these sites if they had genotyping quality score of at least 10 (`--minGQ 10`). I found the vast majority of variants to be shared among all 10 individuals (data not shown), so I applied filters for the maximum and minimum minor allele frequencies, such that these would not be either zero or one (`–maf 0.01 –max-maf 0.99`). Any calls that did not receive a 'PASS' flag were discarded (`--remove-filtered-all`).

@@

```{r overlap_in_sets, engine = 'bash', eval = FALSE, echo = TRUE}
# prepare files for isec by zipping and re-indexing
bgzip mpileup.filter.recode.vcf
bcftools index mpileup.filter.recode.vcf.gz
bgzip freebayes.filter.recode.vcf
bcftools index freebayes.filter.recode.vcf.gz
# Compute intersection of 2 files with isec
bcftools isec -p isec mpileup.filter.recode.vcf.gz freebayes.filter.recode.vcf.gz
# Get overlap data from isec into tsv format
> Overlap.vals
for file in isec/*.vcf
do
  bcftools stats $file | grep "number of SNPs:" | cut -f4 >> Overlap.vals
done
echo $'unique to mpileup\nunique to freebayes\nshared\nshared' > Overlap.names
paste  Overlap.names Overlap.vals > Overlap.tsv
sed -i '1s/^/SNPs\tCount\n/' Overlap.tsv
```

Using `vcftools`, I computed the allele frequency at each loci across individuals, as well as the depth at each locus summed across individuals.

```{r vcf_freq, engine = 'bash', eval = FALSE, echo = TRUE}
# allele frequency
vcftools --gzvcf mpileup.filter.recode.vcf.gz --freq  --out mpileup.filter
vcftools --gzvcf freebayes.filter.recode.vcf.gz --freq --out freebayes.filter
# site depth sum
vcftools --gzvcf mpileup.filter.recode.vcf.gz --site-depth --out mpileup
vcftools --gzvcf freebayes.filter.recode.vcf.gz --site-depth --out freebayes
```

@@

I created figures and summary tables from the data generated above using the `R` language. *This code is presented in the supplementary section at the end of the manuscript.*

@@

# Overlap counts table; add totals for freebayes and mpileup 
Overlap <- read_delim('./data/Overlap.tsv', '\t') %>%  
  distinct() %>% 
  bind_rows(
    tibble(SNPs = c('mpileup', 'freebayes'), 
           Count = c(.[1,2][[1]] + .[3,2][[1]], 
                     .[2,2][[1]] + .[3,2][[1]]))) %>%  
  arrange(desc(Count))
```

```{r load_data2, message=FALSE, echo = F}
# Allele frequencies
cnames <- c('chrom','pos', 'n_alleles','n_chr','a1','a2')
AlleleFreq <-  
  bind_rows(
    read_tsv('./data/freebayes.filter.frq', col_names = cnames, skip = 1L) %>% 
      mutate(caller = 'freebayes'),
    read_tsv('./data/mpileup.filter.frq', col_names = cnames, skip = 1L) %>% 
      mutate(caller = 'mpileup')
  ) %>% 
  mutate(allele = str_extract(a1, '[A-Z]+'), 
         frequency = str_remove_all(a1, '[A-Z]+:'),
         allele2 = str_extract(a2, '[A-Z]+'), 
         frequency2 = str_remove_all(a2, '[A-Z]+:')
  ) %>% 
  mutate(across(contains('freq'), as.numeric)) %>%
  mutate(
    maf = ifelse(frequency > 0.5, 1 - frequency, frequency),
    minor_allele = ifelse(maf == frequency, allele, allele2)
  )

@@

vcf_df <- SiteDepth %>%
  rename(chrom = chr) %>% 
  left_join(AlleleFreq, by = c('chrom', 'pos', 'caller'))

cnames <- c('id', 'quality', 'n_snps', 'ts', 'tv', 'indels')
qual_df <- 
  bind_rows(
    read_tsv("./data/mpileup.quality.stats", skip = 1L, col_names = cnames) %>%  mutate(caller = "mpileup"),
    read_tsv("./data/freebayes.quality.stats", skip = 1L, col_names = cnames) %>%  mutate(caller = "freebayes")
  ) %>%  mutate(caller = as_factor(caller))
```

```{r figs, echo=F}
library(ggbeeswarm)
library(patchwork)
library(rcartocolor)
library(gridExtra)
library(grid)

# Allele frequency histogram (alleles stacked)
fig_allelefreq <- 
  ggplot(AlleleFreq) +
  geom_histogram(aes(x = maf), bins = 20, alpha = 0.8) +
  facet_wrap(caller~., scales = 'free_y') +
  labs(x = 'minor allele frequency', y='loci') +
  theme_bw()

@@

## SUPPLEMENTAL

# Allele depth over maf, coloured by n chromosomes
fig_alleledepth_over_maf <- 
  ggplot(vcf_df, aes(x=maf, y=allele_depth, color = n_chr)) +
  geom_point(size = 0.5, alpha = 0.7) +
  scale_y_log10() +
  facet_wrap(~caller) +
  scale_color_viridis_c() +
  labs(x = "Minor allele frequency",
       y = "Allele depth", 
       color="Number of\nchromosomes") +
  theme_bw()

# The distribution of mean depth per site within individuals
set.seed(45)
fig_indvdepth <-
  ggplot(IndvDepth, aes(y = mean_depth, x = caller)) +
  geom_violin(alpha = 0.1, draw_quantiles = c(.25, .5, .75), trim = F) +
  geom_quasirandom(aes(color = individual), size = 3, width = 0.3) +
  scale_color_carto_d(palette = 2) +
  guides(colour = guide_legend(override.aes = list(size = 4, pch = 16))) +
  facet_wrap( ~ caller, scales = 'free_x') +
  labs(y = 'mean depth per locus', x = '', color = 'Individual') +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = 'right'
  )
fig_quality_dist <- qual_df %>% 
  mutate(QUAL = round(quality/25)*25) %>% 
  group_by(QUAL, caller) %>% 
  summarise(n_snps = sum(n_snps),
            ts  = sum(ts),
            tv = sum(tv)) %>% 
  pivot_longer(cols = c(ts,tv)) %>% 
  ggplot(aes(x = QUAL, y = value, fill = name)) +
  geom_col(width = 19) +
  facet_wrap(~caller) +
  scale_fill_carto_d(palette = 2) +
  labs(x="quality score", y = "loci", fill = NULL) +
  scale_y_log10() +
  theme_bw()
```

@@

The distribution minor allele frequencies of SNPs from `mpileup` shows a U-shape with peaks at 0.05 and 0.5, whereas those from `freebayes` peak at 0.05 and are otherwise evenly distributed (fig.1A). The range of possible values is constrained due to the small sample size (some values can only be obtained with a specific amount of missing data; fig. 2A). The sum depth per locus across all individuals is generally in the range of 50-200 reads, with both callers showing similar distributions (fig. 1B). The mean depth for `mpileup` was 173.7 reads and for `freebayes` 162.2 reads. Both callers have some extreme outliers in terms of site depth: the maximum depth for `freebayes` was 16,162 reads, while for `mpileup` it was only 2522 reads. The minimum site depth was 5 and 3 for `freebayes` and `mpileup` respectively and `mpileup` has a greater proportion of low-depth sites in general. Freebayes also had a greater mean depth per individual (fig. 2B). Both callers show similar distributions of quality scores, though `mpileup` calls a greater proportion of transversions at low-quality sites (fig. 2C).

@@

```{r fig1, echo=F, fig.align = 'center', fig.width=7, fig.height=3, fig.cap="(A) Distribution of minor allele frequency; SNPs with frequency = 0 or 1 were omitted. (B) Distribution of depth per locus summed across all individuals (note log-scaled x-axis)."}
(fig_allelefreq + fig_sitedepth) &
  plot_layout(guides = 'keep') &
  plot_annotation(tag_levels = 'A')
```

```{r table2_overlap, echo=F}
Overlap %>% 
  pander::pander(caption = "Summary of total, shared, and unique SNPs called by mpileup and freebayes (after filtering).")
```

```{r fig2, echo=F, fig.align = 'center', fig.width=6, fig.height=7, fig.cap="(A) Scatterplot of loci (points) by allele depth and minor allele frequency, colored by the number of chromosomes counted. Due to the sample size (n=10), minor allele frequencies are constrained by the number of chromosomes at the SNP site; there appears to be no relationship between allele depth (ie. read coverage) and minor allele frequencies. \n(B) Mean depth per locus for each individual. Violin plot lines indicate quartiles. \n(C) The quality score distributions for each set of SNP calls, separated into transitons (ts) and transversions (tv)"}
fig_alleledepth_over_maf / fig_indvdepth / fig_quality_dist & 
  plot_annotation(tag_levels = 'A')
```

In this work I compared the sets of biallelic SNPs called by two tools, bcftools mpileup/call and freebayes, from alignments of burbot reads from 10 individuals. The raw SNP calls were filtered similarly for both sets based on quality scores, genotyping confidence scores, and minor allele frequency; most of the data (\>90%) was discarded during filtering (table 1), with roughly one-third of calls having allele frequency of 0 or 1 (*ie.* all individuals were similar but differed from the reference). Unexpectedly, I found that mpileup called many more SNPs that freebayes, and retained nearly twice as many after filtering. This is in contrast to findings from @cornish2015 who found that freebayes called more SNPs in human exome data, though both tools had similar-sized sets after filtering. The discrepancy between these findings does not seem to be explained by the slight differences in filtering (they filtered for read depth; I filtered for genotype quality), but could be related to differences between datasets [@hwang2015].

Interestingly, the sets of SNPs called by mpileup and freebayes showed a high degree of overlap, with over 90% of freebayes calls being shared with the larger mpileup set (table 2). This result is somewhat in agreement with @hwang2015, who found \~90% concordance between (Samtools) mpileup, freebayes, and GATK on the gold-standard Genome in a Bottle dataset. My results suggest then, that either mpileup has low specificity (calls variants that don't exist), or that freebayes has low sensitivity (misses true variants). This does not seem to be the case generally though, as @cornish2015 suggest the opposite and found that a variety of tools had similarly low sensitivity on the GiaB gold-standard data.

My analysis also revealed differences in the distribution of minor allele frequencies (MAF) and read depth at SNP loci (fig. 1). Specifically, mpileup shows a clear preference for SNPs with MAF of 0.5 and to a lesser degree 0.25 (relative to similar values). Both callers have a peak at low MAF (0.05) but otherwise mpileup has an even distribution from 0.10 to 0.5, when accounting for the constraint on these values by the small sample size (fig. 2A). The callers show differences in the read depth across loci (fig. 1B) and individuals (fig. 2B), as high-depth sites (\>250/ind.) are hard-filtered by `mpileup`, and mpileup identifies many more low-depth sites (\<70x) that are ignored by freebayes. As such, the concordance between sets would likely increase if stricter filters were used to mitigate these biases.

Certain characteristics of the burbot data present challenges for variant calling. The highly-fragmented assembly means we are likely to miss many variants since they are not represented in the alignment. The relatively low coverage of the data means that a number of true variants are likely discarded because of high uncertainty in their classification. Further, burbot is not a model organism and does not have the same resources to develop useful prior probabilities for variant calling (eg. SNP databases for human variant calling). In the present work, the small number of subjects means that the identification of uncommon and rare variants is highly unlikely. If we were more interested in finding such loci, we would want to apply less stringent filtering and accept the cost of increased false discoveries.

@@

# Supplemental

I also examined the quality score distribution over sites and the mean per site depth by individual. These were computed with `bcftools stats` and `vcftools` respectively, as follows:

```{r engine = 'bash', eval=FALSE}
## quality scores from bf
grep -E "Quality" mpileup.filter.recode.stats | cut -f1 --complement > mpileup.quality.stats
grep ^QUAL mpileup.filter.recode.stats | cut -f1 --complement >> mpileup.quality.stats 
grep -E "Quality" freebayes.filter.recode.stats | cut -f1 --complement > freebayes.quality.stats
grep ^QUAL freebayes.filter.recode.stats | cut -f1 --complement >> freebayes.quality.stats 
# Mean depth for each individual across all loci
vcftools --gzvcf mpileup.filter.recode.vcf.gz --depth --out mpileup
vcftools --gzvcf freebayes.filter.recode.vcf.gz --depth --out freebayes
```

### R code for parsing results data

```{r supplementary1, eval = F, echo=T, include=T}
#! {R code}
library(tidyverse, quietly = T)
library(ggbeeswarm)
library(patchwork)
library(rcartocolor)
library(gridExtra)
library(grid)

# Overlap counts table; add totals for freebayes and mpileup 
Overlap <- read_delim('./data/Overlap.tsv', '\t') %>%  
  distinct() %>% 
  bind_rows(
    tibble(SNPs = c('mpileup', 'freebayes'), 
           Count = c(.[1,2][[1]] + .[3,2][[1]], 
                     .[2,2][[1]] + .[3,2][[1]]))) %>%  
  arrange(desc(Count))

vcf_df <- SiteDepth %>%
  rename(chrom = chr) %>% 
  left_join(AlleleFreq, by = c('chrom', 'pos', 'caller'))

cnames <- c('id', 'quality', 'n_snps', 'ts', 'tv', 'indels')
qual_df <- 
  bind_rows(
    read_tsv("./data/mpileup.quality.stats", skip = 1L, col_names = cnames) %>%  mutate(caller = "mpileup"),
    read_tsv("./data/freebayes.quality.stats", skip = 1L, col_names = cnames) %>%  mutate(caller = "freebayes")
  ) %>%  mutate(caller = as_factor(caller))
```

### R code for plotting results

```{r supplementary2, eval = F, echo=T, include=T}
# Allele frequency histogram (alleles stacked)
fig_allelefreq <- 
  ggplot(AlleleFreq) +
  geom_histogram(aes(x = maf), bins = 20, alpha = 0.8) +
  facet_wrap(caller~., scales = 'free_y') +
  labs(x = 'minor allele frequency', y='loci') +
  theme_bw()

## SUPPLEMENTAL

# Allele depth over maf, coloured by n chromosomes
fig_alleledepth_over_maf <- 
  ggplot(vcf_df, aes(x=maf, y=allele_depth, color = n_chr)) +
  geom_point(size = 0.5, alpha = 0.7) +
  scale_y_log10() +
  facet_wrap(~caller) +
  scale_color_viridis_c() +
  labs(x = "Minor allele frequency",
       y = "Allele depth", 
       color="Number of\nchromosomes") +
  theme_bw()

# The distribution of mean depth per site within individuals
set.seed(45)
fig_indvdepth <-
  ggplot(IndvDepth, aes(y = mean_depth, x = caller)) +
  geom_violin(alpha = 0.1, draw_quantiles = c(.25, .5, .75), trim = F) +
  geom_quasirandom(aes(color = individual), size = 3, width = 0.3) +
  scale_color_carto_d(palette = 2) +
  guides(colour = guide_legend(override.aes = list(size = 4, pch = 16))) +
  facet_wrap( ~ caller, scales = 'free_x') +
  labs(y = 'mean depth per locus', x = '', color = 'Individual') +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = 'right'
  )
fig_quality_dist <- qual_df %>% 
  mutate(QUAL = round(quality/25)*25) %>% 
  group_by(QUAL, caller) %>% 
  summarise(n_snps = sum(n_snps),
            ts  = sum(ts),
            tv = sum(tv)) %>% 
  pivot_longer(cols = c(ts,tv)) %>% 
  ggplot(aes(x = QUAL, y = value, fill = name)) +
  geom_col(width = 19) +
  facet_wrap(~caller) +
  scale_fill_carto_d(palette = 2) +
  labs(x="quality score", y = "loci", fill = NULL) +
  scale_y_log10() +
  theme_bw()
```

### R code to output tables and figures

```{r supplementary3, eval = F, echo=T, include=T}
## table 1
tribble(
  ~ Caller, ~ `raw SNPs`, ~ `raw ts/tv`, ~ `post-filter SNPs`, ~ `post-filter ts/tv`,
  "mpileup", 132046, 0.81, 11115, 0.87,
  "freebayes", 106973, 0.44, 6722, 1.09
) %>% 
  pander::pander(caption = "Raw and filtered SNP counts and transition/transversion ratios for calls by mpileup and freebayes.")

### fig 1
(fig_allelefreq + fig_sitedepth) &
  plot_layout(guides = 'keep') &
  plot_annotation(tag_levels = 'A')

## table 2
Overlap %>% pander::pander(caption = "Summary of total, shared, and unique SNPs called by mpileup and freebayes (after filtering).")

### fig 2
(fig_alleledepth_over_maf / fig_indvdepth / fig_quality_dist) & 
  plot_annotation(tag_levels = 'A')
```
